### 2016-06-13 locking.tex
polishing edits
improve figure placement
add examples for distributed case and third interpretation

@@

While it is intuitively appealing to combine period enforcement with suspension-based locking protocols~\cite{Raj:91,Lak:11,LNR:09}, we observe that this causes non-trivial difficulties. In particular, our examples show that the addition of period enforcement invalidates all existing blocking analyses. They also suggest that devising a correct blocking analysis would be a substantial challenge due to the demonstrated feedback cycle between the period enforcer rules and blocking durations.

Fundamentally, the design of the period enforcer algorithm implicitly rests on the assumption that a segment \emph{can} execute as soon as it is eligible to do so. In the presence of locks, however, this assumption is invalidated. As demonstrated, the result can be a successive growth of self-suspension times that proceeds until a deadline is missed.  The period enforcer algorithm, at least as defined and used in the literature to date~\cite{Raj:suspension1991,Raj:91}, is therefore incompatible with the existing literature on suspension-based real-time locking protocols (e.g., \cite{Raj:91,Lak:11,LNR:09,BLBA:07,Br:13}).

Notably, the examples in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2} assume a shared-memory locking protocol: once a lock is granted, tasks execute their own critical sections, on their assigned processors. One may wonder whether  effects similar to those described in in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2} can also occur under \emph{distributed} real-time locking protocols such as the DPCP~\cite{RSL:88} or the DFLP~\cite{Br:13,Br:14}, where critical sections may be executed on dedicated \emph{synchronization processors}. In this case, the self-suspension occurs on the task's \emph{application processor}, which is different from the (remote) synchronization processor on which the critical section is executed. This separation allows employing period enforcement  only on application processors (while avoiding it on synchronization processors) without incurring the feedback cycle between blocking times and self-suspension times highlighted in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2}. However, period enforcement still invalidates all existing blocking analyses for distributed real-time semaphore protocols~\cite{RSL:88,Raj:91,Br:13} because it artificially increases blocking times if tasks contain multiple accesses to shared resources.

Finally, it is worth noting that our examples can be trivially extended with lower-priority tasks to ensure that no processor idles before the described deadline misses occur. It is also not difficult to extend the second example with a task on a third processor such that all segments of $\tau_1$ and $\tau_2$ are separated by a non-zero self-suspension.

>>>>

\subsection{Other Protocols and Interpretations}

The examples in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2} assume a \emph{shared-memory} locking protocol: once a lock is granted, tasks execute their own critical sections, on their assigned processors. One may wonder whether  effects similar to those described in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2} can also occur under \emph{distributed} real-time locking protocols such as the DPCP~\cite{RSL:88} or the DFLP~\cite{Br:13,Br:14}, where critical sections may be executed on dedicated \emph{synchronization processors}. In this case, the self-suspension occurs on the task's \emph{application processor}, which is different from the (remote) synchronization processor on which the critical section is executed.

This separation allows employing period enforcement  only on application processors (while avoiding it on synchronization processors) without incurring the feedback cycle between blocking times and self-suspension times highlighted in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2}.

However, period enforcement still invalidates all existing blocking analyses for distributed real-time semaphore protocols~\cite{RSL:88,Raj:91,Br:13} because it artificially increases blocking times if tasks contain multiple accesses to shared resources. An example demonstrating this effect is shown in Figure~\ref{fig:locking-dpcp}. Two segmented self-suspending tasks $\tau_1$ and $\tau_2$ share a resource using a distributed real-time locking protocol. The choice of protocol is irrelevant; the example works with both the DPCP and the DFLP. The tasks have parameters $m_1 = 2$, $C_1^1 = C_1^2 = 1$, $T_1 = D_1 = 25$ and $m_2 = 3$, $C_2^1 = C_2^3 = 1$, $C^2_2 = 6$, and $T_2 = D_2 = 15$. The computation segments are separated by self-suspensions that arise while the tasks wait for the completion of critical sections that are executed remotely on a dedicated synchronization processor $P_{sync}$; the corresponding suspension segment parameters $S_1^1$, $S_2^1$, and $S_2^2$ will be defined shortly.

The first jobs of $\tau_1$ and $\tau_2$ are both released at time~0 and attempt to access the shared resource at time~1. Task $\tau_1$'s request is serviced first; as a result $\tau_2$ resumes only at time $a^2_{2,1} = ET_{2,1}^2 = 5$ after having been suspended for four time units:
\begin{align*}
	ET_{2,1}^2 & = \max\left(ET_{2,0}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,1})\right) =\max(-T_2 + T_2,\ 5) = 5.
\end{align*}
Task $\tau_2$ then executes its second computation segment for $C_2^2 = 6$ time units until time~11, when the job accesses the shared resource for a second time. Since there is no contention from $\tau_1$ at this time, $\tau_2$ resumes after only two time units at time~13. This leaves the job sufficient time to complete at time~14, one time unit before its deadline at time~15.

The second job of $\tau_2$ is released at time~15 and issues a request for the shared resource at time~16. Since there is no contention from $\tau_1$ at the time, the second computation segment arrives already at time $a^2_{2,2} = 18$, after having been self-suspended for only two time units. However, since the second segment of the first job arrived at time $a^2_{2,1} = 5$, the second segment of the second job is not eligible to start execution until time $ET_{2,2}^2 = 20$ since 
\begin{align*}
	ET_{2,2}^2 & = \max\left(ET_{2,1}^2 + T_2,\ \mathit{busy}(\tau_2, a^2_{2,2})\right) =\max(5 + T_2,\ 18) = 20.
\end{align*}
As a result, $\tau_2$ faces contention from $\tau_1$ when it issues its second request for the shared resource at time~26, which ultimately leads to a deadline miss at time~30.

In contrast, without period enforcement, $\tau_2$ does not miss its deadline at time~30 because, \emph{across its two requests}, a job of $\tau_2$ is delayed by at most one request of $\tau_1$, for a total self-suspension time of at most two six time units. That is, even though the \emph{individual} self-suspension segments of the two tasks are each up to four time units long (i.e., $S_1^1 = S_2^1 = S_2^2 = 4$), the fact that self-suspensions arise due to the same cause (resource contention) means that the \emph{total} self-suspension time is actually less than the sum of the individual per-segment bounds.

 Existing analyses for the DPCP~\cite{RSL:88,Raj:91,Br:13} and the DFLP~\cite{Br:13} exploit this knowledge and therefore predict task $\tau_2$ to be schedulable with a worst-case response time of 14.\footnote{The analyses in \cite{RSL:88,Raj:91} do assume a segmented task model, but bound the total blocking across all segments. The analysis in \cite{Br:13} also bounds the total blocking across all segments and can be applied to both the segmented and the dynamic self-suspension model.} The example in Figure~\ref{fig:locking-dpcp} thus demonstrates that the period enforcer invalidates existing blocking bounds for distributed semaphore protocols. As an aside, the example in Figure~\ref{fig:locking-dpcp} further highlights limitations of the segmented self-suspension model in the context of synchronization protocols, where the lengths of self-suspensions encountered at runtime are inherently not independent.

\begin{figure}[t]
  \centering
  \includegraphics[scale=1]{../figures/locking3/sched-locking-dpcp.pdf}
  \caption{Example schedule of two tasks $\tau_1$ and $\tau_2$ on two processors sharing a \emph{remote} resource using a distributed semaphore protocol (e.g., the DPCP~\cite{RSL:88,Raj:91} or the DFLP~\cite{Br:13,Br:14}) together with the period enforcer. Since critical sections are executed as \emph{remote procedure calls} (RPCs) on a dedicated synchronization processor $P_{sync}$ (not subject to period enforcement), their execution is not delayed by period enforcement. However, period enforcement delays $\tau_2$ at time~18, which invalidates existing analyses~\cite{RSL:88,Raj:91,Br:13}: under both the DPCP and the DFLP, $\tau_2$ is predicted to have a maximum response time of~14~\cite{Br:13}, but with period enforcement, the second job of $\tau_2$ has in fact a response time of 16.
  }
  \label{fig:locking-dpcp}
\end{figure}

Returning to the shared-memory case,
as a third possible interpretation, one could also exclude critical sections from period enforcement such that only the rest of the computation segment \emph{after} a critical section is subject to period enforcement (i.e., making critical sections immediately eligible to execute).\footnote{This interpretation does not fit the assumptions stated in~\cite{Raj:suspension1991,Raj:91}.} This can be understood as making each critical section an individual computation segment (exempt from period enforcement) that is separated from the following computation by a ``virtual'' self-suspension of maximum length zero. As in the case of distributed semaphore protocols, this interpretation breaks the feedback cycle highlighted in Sections~\ref{sec:sem-case-1} and~\ref{sec:sem-case-2}, but still invalidates all existing blocking analyses as it artificially inflates the synchronization delay.

An example of this effect is shown in  Figure~\ref{fig:locking-alt3}, which depicts the same scenario as in Figure~\ref{fig:locking-dpcp} under the assumption that a shared-memory semaphore protocol is used (i.e., critical sections are executed locally by each job) and that critical sections are exempt from period enforcement. As in the distributed case, period enforcement induces a deadline miss, whereas existing blocking analyses~\cite{Br:13,LNR:09} exploit the fact that a remote critical section can block only once, thus arriving at a worst-case response time bound of 14 for $\tau_2$.

\begin{figure}[t]
  \centering
  \includegraphics[scale=1]{../figures/locking3/sched-locking-alt3.pdf}
  \caption{Example schedule of two tasks $\tau_1$ and $\tau_2$ on two processors sharing one lock-protected resource. The example assumes that lock requests take effect immediately and that critical sections are exempt from period enforcement (i.e., period enforcement is applied only to any computation after a critical section). As in Figure~\ref{fig:locking-dpcp}, the second job of task $\tau_2$ misses its deadline at time $30$.
  }
  \label{fig:locking-alt3}
\end{figure}

To conclude, in both Figures~\ref{fig:locking-dpcp} and~\ref{fig:locking-alt3}, period enforcement has an effect  \emph{as if} a single remote critical section can block a given job multiple times, which is fundamentally incompatible with efficient blocking analysis~\cite{Br:13}.

@@

While it is intuitively appealing to combine period enforcement with suspension-based locking protocols~\cite{Raj:91,Lak:11,LNR:09}, we observe that this causes non-trivial difficulties. In particular, our examples show that the addition of period enforcement invalidates all existing blocking analyses.

If critical sections are subject to period enforcement, our examples also suggest that devising a correct blocking analysis would be a substantial challenge due to the demonstrated feedback cycle between the period enforcer rules and blocking durations. Fundamentally, the design of the period enforcer algorithm implicitly rests on the assumption that a segment \emph{can} execute as soon as it is eligible to do so. In the presence of locks, however, this assumption is invalidated. As demonstrated, the result can be a successive growth of self-suspension times that proceeds until a deadline is missed.  The period enforcer algorithm, at least as defined and used in the literature to date~\cite{Raj:suspension1991,Raj:91}, is therefore incompatible with the existing literature on suspension-based real-time locking protocols (e.g., \cite{Raj:91,Lak:11,LNR:09,BLBA:07,Br:13}).

Finally, it is worth noting that our examples can be trivially extended with lower-priority tasks to ensure that no processor idles before the described deadline misses occur. It is also not difficult to extend the examples in Figures~\ref{fig:locking-alt2} and~\ref{fig:locking-alt3} with a task on a third processor such that all critical sections of $\tau_1$ and $\tau_2$ are separated from their predecessor segments by a non-zero self-suspension.
